A fast algorithm for global code motion of congruent instructions

### 2018-02-23 cgo/paper.tex
cc comments

@@

\def \GCC {GCC}
\def \LLVM {LLVM}
\def \CFG {CFG}
\def \SSA {SSA}
\def \MemorySSA {MemorySSA}
\def \PRE {PRE}
\def \GVN {GVN}
\def \SPEC {SPEC Cpu 2006}
\def \gcm {global-code-motion}
\def \GCM {GCM}
\def \xlinux {x86\_64-linux}
\def \ooo {out-of-order}

@@

\subsection{Barriers}
\label{subsec:barriers}
Barriers are based on the concept of pinned instructions \cite{click1995global}
but extended to adapt to \LLVM{} IR. A basic-block in \LLVM{} IR is actually an
extended basic-block because there might be non-returning calls in the middle of
a basic-block. Essentially, any instruction that cannot guarantee progress is
marked as a barrier. A barrier will prevent any instruction occurring lexically
below it from getting hoisted. On the other hand, a sink barrier will prevent
any instruction above it from getting sunk. If there are no barriers in a basic
block, any instruction satisfying other legality checks can be sunk.

\begin{lstlisting}
// Compute both hoist and sink barriers
void computeBarriers()
  for each basic-block B in a function:
    barrier_found = false
    for each instruction I in B:
      if I does not guarantee progress:
        mark I as a barrier instruction
        barrier_found = true
        break;

@@

\begin{lstlisting}
B0: i0 = load B
B1: i1 = load A
    c1 = i1 + 10
    d1 = i0 + 20
    goto B3
B2: i2 = load A
    c2 = i2 + 10
    d2 = i0 + 20
    goto B3

@@

In this example (c1, c2) or (d1, d2) are potential sinkable candidates. Since
(c1, c2) depend on i1 and i2 respectively which are also in their original basic
blocks, c1 and c2 are not fully anticipable in B3. So without knowing the
ability to sink `i1' and `i2' it would be illegal to sink (c1, c2) to B3. On the
other hand (d1, d2) can safely be sunk because their operands are readily
available at the sink point, i.e., B3. It should also be noted that, just
because the expressions are identical and operands are available, it still
requires a unique post-dominating $\Phi$ to use the exact same values to be
legally sinkable.

\subsection{Profitability check (Cost models)}
\label{subsec:cost-models}
After the legality checks have passed, we check if a \GCM{} is profitable.  That
takes into account the impact \GCM{} would have on various parameters that might
affect runtime performance, e.g., impact on live-range, gain in the code
size. The current implementation makes effort to not regress in performance but
reduce register spills. We implement a cost model to reduce register pressure.

\subsubsection{Reduce register pressure}
\label{hoist:reg-pressure}
Hoisting upwards will decrease the live-range of its use, if it is a last use (a
kill) but increase the live-range of its definition. Conversely, sinking will
decrease the live-range of the defined register but increase the live-range for
killed operands. If the live-range after \GCM{} will decrease, it will be
moved. Essentially, as long as there is one killed operand, code hoisting will
either decrease or preserve the register pressure.  Similarly, code-sinking will
either decrease or preserve the register pressure as long as there is one
operand killed at most.  The following example explains how \GCM{} of can reduce
the register pressure.

@@

\subsubsection{Hoisting an expression across a call}
\label{cost:across-calls}
Call expressions are difficult to analyze in an intraprocedural optimization
pass like \GCM{}. Even hoisting scalars across calls is tricky because it can
increase the number of spills. During the register allocation of calls
expressions, the argument registers (also known as the caller saved registers)
are saved because they might be modified by the callee and after the call they
are restored. Due to this the register pressure is high because the number of
available registers are reduced by the number of caller saved registers. In that
situation a computation that reduces register pressure may be beneficial to
hoist. Similarly, a computation that introduces a register (without killing any
other operand) before a call may be beneficial to sink.

\section{A fast algorithm for code motion of congruent computations}
\label{subsec:optimistic}
Once all the legality and profitability checks are satisfied for a set of
congruent instructions, they are suitable candidates for code motion. A copy of
the computation is inserted at the hoisting/sinking point along with any
instructions which needed to be rematerialized. Thereafter, all the computations
made redundant by the new copy are removed and affected data structuers are
updated.

\begin{lstlisting}
void doGCM()
  Analyses available:
    Dominator Tree, DFS Numbering, Memory SSA
  Compute DJ-graph of function
  constructMergeSet(CFG)
  computeBarriers()
  do:
    Compute GVN of each expression in the function
    // Repeat hoisting if any candidates were hoisted.
  while (doCodeHoisting() > 0)
  // Code-sinking only once
  doCodeSinking()
\end{lstlisting}

@@

\begin{lstlisting}
int doCodeHoisting(VNs)
  // VNs = map (VN -> list of I with value VN)
  Sort VNs in ascending order

  for each VN in VNs
    for each instruction with the same VN
      Find its post-dominance-frontier (PDF)
      Insert the CHI-node at PDF

@@

\begin{lstlisting}
int doCodeSinking()
  // VNs = map (VN -> list of I with value VN)
  Sort VNs in ascending order

  for each VN in VNs
    for each instruction with the same VN
      Find its dominance-frontier (DF)
      Insert the $\Phi$-node at DF

@@

\subsection{Time complexity of algorithm}
The complexity of code hoisting is linear in number of instructions that are
candidates for \gcm{}, matching the complexity of \PRE{} on \SSA{} form.  The
analyses computed for this pass like Global Value Numbering, Marking Barriers,
are both linear in number of instructions in a function. Computation of DJ-graph
and merge-sets is $O(E)$, E being number of edges in the CFG. Liveness analysis
is not very expensive \cite{das2012} but still only performed on-demand for
hoistable candidates. Other analyses like Alias Analysis, \MemorySSA{} and
Dominator Tree are already available in the \LLVM{} pass pipeline.

@@

\section{Experimental Evaluation}
\label{sec:experimental-results}
% Experimental results: Register spills, compile time, run time. compile-time comparison w.r.t. gcc
For evaluation of \gcm{}, we built \SPEC{} with and without the \GCM{}
optimization. All the experiments were conducted on an \xlinux{} machine and at
-Ofast optimization level with inlining and loop unrolling disabled; inliner and
loop-unroller vary wildly with \GCM{} optimization so it was difficult to
visualize the impact. Each benchmark was run three times and the best result was
taken. We collected execution time and code-size which is listed in
Table~\ref{tab:code-size}, other compiler statistics are listed in
Table~\ref{tab:stats}, Table~\ref{tab:spills}, and
Figure~\ref{tab:code-motion-metric}.  We also measured compile time
Table~\ref{tab:compile-time}. The benchmark 464.h264ref caused an internal
compiler error with the trunk clang so it could not be reported.

>>>>

\renewcommand\lstlistingname{Algorithm}

@@

\def \GCC {GCC}
\def \LLVM {LLVM}
\def \CFG {CFG}
\def \SSA {SSA}
\def \MemorySSA {MemorySSA}
\def \PRE {PRE}
\def \GVN {GVN}
\def \SPEC {SPEC CPU 2006}
\def \gcm {global-code-motion}
\def \GCM {GCM}
\def \xlinux {x86\_64-linux}
\def \ooo {out-of-order}

@@

\subsection{Barriers}
\label{subsec:barriers}
Barriers are based on the concept of pinned instructions \cite{click1995global}
but extended to adapt to \LLVM{} IR. A basic-block in \LLVM{} IR is actually an
extended basic-block \cite{ebb} because there might be non-returning calls in
the middle of a basic-block. There might be other instructions like stores,
where compiler may not guarantee progress of control flow, all such instructions
are marked as barriers. A barrier will prevent any instruction occurring
lexically below it from getting hoisted. On the other hand, a sink barrier will
prevent any instruction above it from getting sunk. If there are no barriers in
a basic block, any instruction satisfying other legality checks can be sunk.

\begin{lstlisting}[caption={Compute hoist and sink barriers},captionpos=b]
// Compute both hoist and sink barriers
void computeBarriers()
  for each basic-block B in a function:
    barrier_found = false
    for each instruction I in B:
      if I does not guarantee progress:
        mark I as a barrier instruction
        barrier_found = true
        break;

@@

\begin{lstlisting}
B0: i0 = load B
    if p is true then goto B1 else goto B2
B1: i1 = load A
    c1 = i1 + 10
    d1 = i0 + 20
    goto B3
B2: i2 = load A
    c2 = i2 + 10
    d2 = i0 + 20
    goto B3

@@

In this example (c1, c2) or (d1, d2) are potential sinkable candidates.  (c1,
c2) depend on i1 and i2 respectively which are also in their original basic
blocks; c1 and c2 are not fully anticipable in B3. So without knowing the
ability to sink `i1' and `i2' it would be illegal to sink (c1, c2) to B3. On the
other hand (d1, d2) can safely be sunk because their operands are readily
available at the sink point, i.e., B3. It should also be noted that, even if the
expressions are identical and operands are available, it still requires a both
instructions to be post-dominated by a unique $\Phi$ to be legally sinkable.

\subsection{Profitability check (Cost models)}
\label{subsec:cost-models}
After the legality checks have passed, we check if a \GCM{} is profitable.  That
takes into account the impact \GCM{} would have on various parameters that might
affect runtime performance, e.g., impact on live-range, gain in the code
size. The current implementation makes effort to not regress in performance and
still reduce register spills. We implement a cost model to reduce register
pressure.

\subsubsection{Reduce register pressure}
\label{hoist:reg-pressure}
Hoisting upwards will decrease the live-range of its use, if it is a last use (a
kill) but increase the live-range of its definition. Conversely, sinking will
decrease the live-range of the defined register but increase the live-range for
killed operands. If the live-range after \GCM{} will decrease, it will be
moved. Essentially, if there is one killed operand, code hoisting will either
decrease or preserve the register pressure as long as the hoisting of killed
operands does not cross a use.  Similarly, code-sinking (of an instructions
which defines a register) will either decrease or preserve the register pressure
as long as there is one operand killed at most and the instruction is not sunk
below the killing use of a non-killed operand. The following example explains
how \GCM{} can reduce the register pressure.

@@

\subsubsection{Hoisting an expression across a call}
\label{cost:across-calls}
Call expressions are difficult to analyze in an intraprocedural optimization
pass like \GCM{}. Even hoisting scalars across calls is tricky because it can
increase the number of spills. During the register allocation of call
expressions, the argument registers (also known as the caller saved registers)
are saved because they might be modified by the callee and after the call they
are restored. Due to this frame setup the register pressure is high because the
number of available registers are reduced by the number of caller saved
registers. In that situation a computation that reduces register pressure may be
beneficial to hoist. Similarly, a computation that introduces a register
(without killing any other operand) before a call may be beneficial to sink.

\section{A fast algorithm for code motion of congruent computations}
\label{subsec:optimistic}
Once all the legality and profitability checks are satisfied for a set of
congruent instructions, they are suitable candidates for code motion. A copy of
the computation is inserted at the hoisting/sinking point along with any
instructions which needed to be rematerialized. Thereafter, all the computations
made redundant by the new copy are removed and affected data structures are
updated.

\begin{lstlisting}[caption={Global Code Motion},captionpos=b]
void doGCM()
  Analyses available:
    Dominator Tree, DFS Numbering, Memory SSA
  Compute DJ-graph of function
  constructMergeSet(CFG)
  computeBarriers()
  do:
    Compute GVN of each expression in the function
    // Repeat hoisting if any candidates were hoisted.
  while (doCodeHoisting() > 0)
  // Code-sinking only once
  doCodeSinking()
\end{lstlisting}

@@

\begin{lstlisting}[caption={Hoist instructions},captionpos=b]
int doCodeHoisting(VNs)
  // VNs = map (VN -> list of I with value VN)
  Sort VNs in ascending order

  for each VN in VNs:
    for each instruction with the same VN:
      Find its post-dominance-frontier (PDF)
      Insert the CHI-node at PDF

@@

\begin{lstlisting}[caption={Sink instructions},captionpos=b]
int doCodeSinking()
  // VNs = map (VN -> list of I with value VN)
  Sort VNs in ascending order

  for each VN in VNs:
    for each instruction with the same VN:
      Find its dominance-frontier (DF)
      Insert the $\Phi$-node at DF

@@

\subsection{Time complexity of algorithm}
The complexity of code hoisting is linear in the number of instructions that are
candidates for \gcm{}, matching the complexity of \PRE{} on \SSA{} form.  The
analyses computed for this pass like Global Value Numbering, Marking Barriers,
are both linear in number of instructions in a function. Computation of DJ-graph
and merge-sets is $O(E)$, E being number of edges in the CFG. Liveness analysis
is not very expensive \cite{das2012} but still only performed on-demand for
hoistable candidates. Other analyses like Alias Analysis, \MemorySSA{} and
Dominator Tree are already available in the \LLVM{} pass pipeline.

@@

\section{Experimental Evaluation}
\label{sec:experimental-results}
% Experimental results: Register spills, compile time, run time. compile-time comparison w.r.t. gcc
For evaluation of \gcm{}, we built \SPEC{} with and without the \GCM{}
optimization. All the experiments were conducted on an \xlinux{} machine and at
-Ofast optimization level with inlining and loop unrolling disabled; inliner and
loop-unroller vary wildly with \GCM{} optimization so it was difficult to
visualize the impact. Each benchmark was run three times and the best result was
taken. We collected execution time and code-size which is listed in
Table~\ref{tab:code-size}, other compiler statistics are listed in
Table~\ref{tab:stats}, Table~\ref{tab:spills}, and
Figure~\ref{tab:code-motion-metric}.  We also measured compile time as shown in
Table~\ref{tab:compile-time}. The benchmark 464.h264ref caused an internal
compiler error with the trunk clang so it could not be reported.
